### 2017-04-25 paper/jfp-submission/safe.tex
reorganize safe.tex

An immediate question arises regarding that 13--16\% of programs for
which we could not synthesize a witness:
%
are they \emph{actually} safe (\ie is the type system being too conservative),
%
or did \toolname simply fail to find the witness?

\ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%

\paragraph{Overloaded Functions.}
%
We found that for 247 programs (roughly a third of the failures, or 5\%
of all programs), \toolname got stuck when it tried to compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

@@

\paragraph{} \hspace{0ex} \\
The remaining 473 programs for which we could not produce a witness did
not admit such an automatic diagnosis, so we selected a random sample of
50 programs and manually searched for a witness.
%
We categorized the programs into three groups.

\paragraph{Non-Parametric Function Type.}
%
44\% of the samples lack a witness in our
semantics due to our non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code.}
%
40\% of the samples contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

12\% of the samples contained type errors that were
unreachable by any inputs, often due to overlapping patterns in a
@match@ expression.
%
% While technically safe, dead code is generally considered a ``code
% smell''; students would likely benefit from a warning in this case.
% \ES{Wes says this is too informal, source the claim}
%
While technically safe, dead code is generally considered a maintenance
risk, as the programmer may not realize that it is dead~\cite{Wheeler2014-fg}
or may accidentally bring it back to life~\cite{Seven2014-gf}.
%
Thus, a warning like that provided by \ocaml's pattern-exhaustiveness
checker would be helpful.

%\paragraph{Bad Function Calls}
A further 28\% of the samples included a function call where the
student had supplied ill-typed inputs, yet the path induced
by the call contained no errors.
%
Consider, for example the following |assoc| function, which
looks up a key in an \emph{association list} and returns a default if
the key cannot be found.
%
\begin{code}
  let rec assoc (d, k, l) = match l with
    | (ki, vi)::tl ->
       if ki = k then
         vi
       else
         assoc (d, k, tl)
    | _ -> d

@@

\paragraph{Witness Exists.}
%
We found that 16\% of the samples admitted a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.

>>>>

An immediate question arises regarding the 13--16\% of programs for
which we could not synthesize a witness:
%
are they \emph{actually} safe (\ie is the type system being too conservative),
%
or did \toolname simply fail to find the witness?
%
% \ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We used a combination of automatic and manual coding to categorize these
programs into four classes.
%
The first class is easily detected by \toolname itself, and thus admits
a precise count.
%
This left us with 473 programs requiring manual coding; we selected
a random sample of 50 programs to investigate, and will report results
based on that sample.
%
Figure~\ref{fig:no-witness} summarizes the results of our investigation,
we note the classes that were based on a random sample with a ``*''.

\begin{figure}[t]
%\centering
\includegraphics[width=0.8\linewidth]{distrib_ext.png}
\vspace{-0.75cm}
\caption{Results of our investigation into programs where \toolname
  did not produce a witness. A ``*'' denotes that the percentage is an
  estimate based on a random sampling of 50 programs.}
\label{fig:no-witness}
\end{figure}

\paragraph{Overloaded Functions.}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

@@

% \paragraph{} \hspace{0ex} \\
% The remaining 473 programs for which we could not produce a witness did
% not admit such an automatic diagnosis, so we selected a random sample of
% 50 programs and manually searched for a witness.
% %
% We categorized the programs into three groups.

\paragraph{Non-Parametric Function Type *.}
%
5\% of programs lack a witness in our semantics due to our
non-parametric $\tfun$ type for functions.
%
Recall that our goal is to expose the runtime errors that would have
been prevented by the type systems.
%
At runtime, it is always safe to call a function, thus we give functions
a simple type $\tfun$ that says they may be applied, but says nothing
about their inputs or outputs.
%
But consider the following @clone@ function, which is supposed to
produce a list containing @n@ copies of the input @x@.
%
% Seven programs violated the \emph{occurs check} with cyclic typing
% constraints like @'a = 'a list@, for example the following @clone@
% function.
%
\begin{code}
  let rec clone x n =
    if n > 0 then
      clone [x] (n - 1)
    else
      []
\end{code}
%
Unfortunately, the student instead constructs an @n@-level nested list
containing a single @x@.
%
The \ocaml compiler rejects this program because the recursive call to
@clone@ induces a cyclic typing constraint @'a = 'a list@, capturing the
fact that each call increases the nesting of the list.
%
\toolname fails to catch this because we do not track the types of the
inputs to @clone@.

@@

\paragraph{Dead Code and ``Safe'' Function Calls *.}
%
4\% of programs contained type errors that were unreachable, either
because they were dead code, or because the student called the function
with inputs that could not trigger the error.

1\% contained type errors that were unreachable by any inputs, often due
to overlapping patterns in a @match@ expression.
%
% While technically safe, dead code is generally considered a ``code
% smell''; students would likely benefit from a warning in this case.
% \ES{Wes says this is too informal, source the claim}
%
While technically safe, dead code is generally considered a maintenance
risk, as the programmer may not realize that it is dead~\cite{Wheeler2014-fg}
or may accidentally bring it back to life~\cite{Seven2014-gf}.
%
Thus, a warning like that provided by \ocaml's pattern-exhaustiveness
checker would be helpful.

%\paragraph{Bad Function Calls}
A further 3\% included a function call where the
student had supplied ill-typed inputs, yet the path induced
by the call contained no errors.
%
Consider, for example the following |assoc| function, which
looks up a key in an \emph{association list} and returns a default if
the key cannot be found.
%
\begin{code}
  let rec assoc (d, k, l) = match l with
    | (ki, vi)::tl ->
       if ki = k then
         vi
       else
         assoc (d, k, tl)
    | _ -> d

@@

\paragraph{Witness Exists *.}
%
We found that only 2\% of programs admit a witness that \toolname
was unable to discover.
%
Slightly over half involved synthesizing a \emph{pair} of
specially-crafted inputs that would result in the function returning
values of incompatible types.
%
The rest required synthesizing an input that would trigger a
particular path through the program, and would likely have been caught
by symbolic execution.
